---
title: "Homework 2"
author: "Your Name"
date: "Due @ 5pm on September 29, 2017"
output:
  pdf_document: default
  html_document: default
header-includes:
- \usepackage{bm}
- \newcommand{\Real}{\mathbb{R}}
- \newcommand{\dom}{{\bf dom}\,}
- \newcommand{\Tra}{^{\sf T}} % Transpose
- \newcommand{\Inv}{^{-1}} % Inverse
- \def\vec{\mathop{\rm vec}\nolimits}
- \def\sweep{\mathop{\rm sweep}\nolimits}
- \newcommand{\diag}{\mathop{\rm diag}\nolimits}
- \newcommand{\tr}{\operatorname{tr}} % Trace
- \newcommand{\epi}{\operatorname{epi}} % epigraph
- \newcommand{\V}[1]{{\bm{\mathbf{\MakeLowercase{#1}}}}} % vector
- \newcommand{\VE}[2]{\MakeLowercase{#1}_{#2}} % vector element
- \newcommand{\Vn}[2]{\V{#1}^{(#2)}} % n-th vector
- \newcommand{\Vtilde}[1]{{\bm{\tilde \mathbf{\MakeLowercase{#1}}}}} % vector
- \newcommand{\Vhat}[1]{{\bm{\hat \mathbf{\MakeLowercase{#1}}}}} % vector
- \newcommand{\VtildeE}[2]{\tilde{\MakeLowercase{#1}}_{#2}} % vector element
- \newcommand{\M}[1]{{\bm{\mathbf{\MakeUppercase{#1}}}}} % matrix
- \newcommand{\ME}[2]{\MakeLowercase{#1}_{#2}} % matrix element
- \newcommand{\Mtilde}[1]{{\bm{\tilde \mathbf{\MakeUppercase{#1}}}}} % matrix
- \newcommand{\Mhat}[1]{{\bm{\hat \mathbf{\MakeUppercase{#1}}}}} % matrix
- \newcommand{\Mcheck}[1]{{\bm{\check \mathbf{\MakeUppercase{#1}}}}} % matrix
- \newcommand{\Mbar}[1]{{\bm{\bar \mathbf{\MakeUppercase{#1}}}}} % matrix
- \newcommand{\Mn}[2]{\M{#1}^{(#2)}} % n-th matrix
---

**Part 1.** 

1. Let $\M{A} \in \Real^{m \times m}$ be a $k$-banded matrix, namely $\ME{A}{ij} = 0$ if $\lvert i - j \rvert > k$.

```{r}
set.seed(12345)
k <- 3; m <- 10
A <- matrix(rnorm(m**2), m, m)
A <- A + t(A)
A <- round(10*A)/10
for (i in 1:m) {
  for (j in 1:m) {
    if (abs(i - j) > k) A[i,j] <- 0
  }
}
A
```

The R snippet above constructs a `r k`-banded `r m`-by-`r m` matrix $\M{A}$. Suppose further that $\M{A}$ is positive definite and therefore admits a unique Cholesky decomposition $\M{A} = \M{R}\Tra\M{R}$ where $\M{R} \in \Real^{m \times m}$ is upper triangular.

Prove that the banded Cholesky decomposition can be computed in $\mathcal{O}(nk^2)$ flops.

**Answer:**

Write your answer here.

Proof:

Step 1: we want to do decomposition like this.

$$\begin{pmatrix}\alpha & \V{a}^T\\\V{a} & \M{A}_* \end{pmatrix}=\begin{pmatrix}\rho & \V{0}^T\\ \V{r} & \M{R}_*\end{pmatrix} 
\begin{pmatrix}\rho & \V{r}^T\\ \V{0} & \M{R}_*^T\end{pmatrix}$$

$$\begin{aligned}
\rho&=\sqrt{\alpha}&&1 flop\\
\V{r}&=\rho^{-1}\V{a}=(a_1,a_2,\cdots,a_{n-1})\cdot\frac{1}{\rho}=(a_1,a_2,\cdots,a_k,0,0,\cdots,0)\cdot\frac{1}{\rho}&&k flops\\
\M{R}_*^T\M{R_*}&=\M{A}_*-\V{r}\V{r}^T=\begin{pmatrix}
a_{*{1,1}} &\cdots & a_{*1,k} & 0 & \cdots & 0\\
\vdots&\ddots&\vdots&\ddots&\ddots&\vdots\\
a_{*k,1}&\cdots&a_{*k,k}&a_{*k,k+1}&\cdots&0\\
a_{*k+1,1}&\cdots&a_{*k+1,k}&a_{*k+1,k+1}&\ddots&0\\
\vdots&\ddots&\vdots&\vdots&\ddots&\vdots\\
0&\cdots&a_{*n-1,n-k}&a_{*n-1,n-k+1}&\cdots&a_{*n-1,n-1}\\
\end{pmatrix}-
\begin{pmatrix}
r_1r_1&\cdots&r_1r_k&0&\cdots&0\\
\vdots&\ddots&\vdots&\vdots&\ddots&0\\
r_kr_1&\cdots&r_kr_k&0&\cdots&0\\
0&\cdots&0&0&\cdots&0\\
\vdots&\ddots&\vdots&\vdots&\ddots&\vdots\\
0&\cdots&0&0&\cdots&0
\end{pmatrix}&&2k^2 flops
\end{aligned}$$

Thus, step 1 needs $(2k^2+k+1)$ flops.

From step 1 to step $n-k-1$, each step needs $(2k^2+k+1)$ flops.

Step $n-k$:
$$\begin{aligned}
\rho&=\sqrt{\alpha}&&1 flop\\
\V{r}&=\rho^{-1}\V{a}=(a_1,a_2,\cdots,a_{k-1})\cdot\frac{1}{\rho}&&k-1  flops\\
\M{R}_*^T\M{R_*}&=\M{A}_*-\V{r}\V{r}^T=\begin{pmatrix}
a_{*{1,1}} & \cdots & a_{*1,k-1}\\
a_{*2,1}   & \cdots & a_{*2,k-1}\\
\vdots&\ddots&\vdots\\
a_{*k-1,1}&\cdots&a_{*k-1,k-1}\\
\end{pmatrix}-
\begin{pmatrix}
r_1r_1&\cdots&r_1r_{k-1}\\
\vdots&\ddots&\vdots\\
r_{k-1}r_1&\cdots&r_{k-1}r_{k-1}\\
\end{pmatrix}&&2(k-1)^2 flops
\end{aligned}$$

Thus, step $n-k$ needs $(2(k-1)^2+(k-1)+1)$ flops.

For step $i$, when $n-k \leq i \leq n$, it needs $(2(n-1-i)^2+(n-1-i)+1)$ flops.

Therefore, all the n steps need $(2k^2+k+1)\cdot(n-k-1)+\sum_{i=n-k}^{i=n}(2(n-i-1)^2+(n-1-i)+1)$ flops, which is $O(k^2n+k^3)$. Since $k \leq n$, it needs $O(nk^2)$ flops.





2. Prove that solving a linear system with the Cholesky factor matrix $\M{R}$ of a $k$-banded matrix $\M{A}$ is $\mathcal{O}(nk)$ flops.

**Answer:**

Write your answer here.
 Proof:
 $$
 \begin{aligned}
 \M{A}\V{x}&=&\V{y}\\
 \M{R^TR}\V{x}&=&\V{y}\\
 \end{aligned}
 $$

According to Problem 1, from step 1 to step $(n-k-1)$, the first k elements in $\V{r}$ calculated in each step may be nonzero while the rest elements must be zero. From step $(n-k)$ to step $n$, all the elements in $\V{r}$ in each step maybe nonzero. Therefore,$$\M{R}=\begin{pmatrix}r_{1,1} & r_{1,2} &\cdots & r_{1,k+1}  & 0 &0&\cdots & 0\\
  0     & r_{2,2} &\cdots & r_{2,k+1} & r_{2,k+2} & 0& \cdots & 0&\\
  \vdots&0&r_{3,3}&\cdots&r_{3,k+2}&r_{3,k+3}&\cdots&0&\\
  0 &\cdots &0&r_{n-k,k+1}&r_{n-k,k+2}&r_{n-k,k+3}&\cdots&r_{n-k,n}\\
  0 &\cdots & 0 & 0 &r_{n-k+1,k+2}&r_{n-k+1,k+3}&\cdots&r_{n-k+1,n}\\
  0 & 0 & \cdots & 0 & 0 &r_{n-k+2,k+3}&\ddots &r_{n-k+2,n}\\
  \vdots&\ddots &\ddots &\ddots &\ddots &0&\ddots&\vdots\\
  0&\cdots&\cdots&\cdots&\cdots&\cdots&\cdots&r_{n,n}
  \end{pmatrix}$$

Step 1: To get $\M{R}\V{x}$, you have to solve $\M{R}^T\V{\beta}=\V{y}$,it needs $O(nk)$ flops;

$$\begin{aligned}
\beta_1&=\frac{y_1}{r_{11}}&&1 flop\\
\beta_2&=\frac{y_2-r_{12}\beta_1}{r_{22}}&&3 flops\\
\vdots&&&\\
\beta_{k+1}&=\frac{y_k-r_{1k+1}\beta_1-r_{2k+1}\beta_2-\cdots-r_{k-1 k+1}\beta_{k-1}}{r_{k+1k
+1}}&& 2k+1 flops\\
\vdots&&&\\
\beta_n&=\frac{y_n-r_{n-kn}\beta_{n-k}-r_{n-k+1n}\beta_{n-k+1}-\cdots-r_{nn}\beta_n}{r_{nn}}&&2k+1 flops
\end{aligned}$$

Thus,it needs $\sum_{i=1}^{i=k}(2i+1)+(2k+1)(n-k)$ flops, which is $O(nk)$ flops.

Step 2: To get $\V{x}$, the situation is the same as step 1. Thus, it needs $O(nk)$ flops.

Therefore, the total is $O(nk)$ flops.

\newpage

**Part 2.** Banded Cholesky Decomposition

You will next add an implementation of the banded Cholesky decomposition to your R package.

Please complete the following steps.

**Step 0:** Make a file called `cholesky_banded.R` in your R package. Put it in the R subdirectory, namely we should be able to see the file at github.ncsu.edu/unityidST758/unityidST758/R/cholesky_banded.R

**Step 1:** Write a function `check_k_banded` that returns `True` if an input matrix `A` is $k$-banded and `False` otherwise. It should return an error message if $k < 0$ or if $k$ is not an integer. Please use the `stop` function. You may also find the function `is.integer` helpful.



**Step 2:** Write a recursive function `chol_banded` that computes the $k$-banded Cholesky decomposition.




- Use sparse matrices via `Matrix` package. Be sure to put the line
```
Depends: Matrix
```
in your `DESCRIPTION` file.
- Your function should return an error message if the input matrix is not $k$-banded (**Hint:** Use your `check_k_banded` function you wrote in Step 1).
- Your function should return an error message if the input matrix is not symmetric
- Your function should return a sparse uppertriangular matrix `R`.

**Step 3:** Write a unit test function `test-cholesky` that

- checks the error messages for `check_k_banded`
- checks the correctness of your `check_k_banded` function
- checks the error messages for `chol_banded`
- checks the correctness of your `chol_banded` output matrix with that obtained by R's `chol` function.

**Step 4:** Apply your `chol_banded` function on $k$-banded positive definite matrices of sizes $n = 25, 50, 100, 200, 400$ and $k = 2, 4, 8$, and plot the run times using `system.out`. Plot runtime versus $n$ and overlay the curves for $k = 2, 4, 8$. Does the run time scale as you'd expect?

```{r}
library("ywang225ST758")

generate_k_banded_pd<-function(n,k){
  R=matrix(numeric(n^2),ncol=n,nrow=n)
  for(i in 1:n){
    R[i,i:min(i+k,n)]=rep(1,min(k+1,n-i+1))
  }
  A=t(R)%*%R+2*diag(rep(1,n),nrow=n,ncol = n)
  return(A)
}


n_vector_step4=c(25,50,100,200,400)
k_vector_step4=c(2,4,8)

time_matrix_step4=matrix(data=numeric(length(n_vector_step4)*length(k_vector_step4)),
                   nrow = length(k_vector_step4),ncol=length(n_vector_step4))

colnames(time_matrix_step4)=as.character(n_vector_step4)
rownames(time_matrix_step4)=as.character(k_vector_step4)

for(i in 1:length(n_vector_step4)) 
  for(j in 1:length(k_vector_step4))
  {
    n=n_vector_step4[i]; k=k_vector_step4[j];
    y=1:n;
    
    A=generate_k_banded_pd(n,k)
    print(paste0("n=",n," k=",k))
    time=system.time(chol_banded(A,k))
    time_matrix_step4[j,i]=time[1] # i for n; j for k
    print(time[1])
  }



j=1
plot(n_vector_step4,time_matrix_step4[j,],xlab="n",ylab="time",ylim=c(0,2*time_matrix_step4[j,5]),
     main="Relation Between time and n,k in step 4")
lines(n_vector_step4,time_matrix_step4[j,])
j=2
points(n_vector_step4,time_matrix_step4[j,])
lines(n_vector_step4,time_matrix_step4[j,],col="red")
j=3
points(n_vector_step4,time_matrix_step4[j,])
lines(n_vector_step4,time_matrix_step4[j,],col="blue") 


legend("topleft",legend = paste0("k=",k_vector_step4),col=c("black","red","blue","green"),
       lty=rep(1,4) )

```

**Step 5:** Generate random right hand side vectors $\V{y}$ of sizes $n = 25, 50, 100, 200,$ and $400$. Using the Cholesky factor matrices $R$ obtained for $k$-banded positive definite matrices of sizes $n = 25, 50, 100, 200, 400$ and $k = 2, 4, 8$, plot the run times solving the linear system $\M{R}\Tra\V{u} = \V{y}$ plus the run time of solving the linear system $\M{R}\V{x} = \V{u}$ as a function of $n$ for $k = 2, 4, 8$. Use the function `Matrix::solve` with the parameter `system` = `"L"` and `"Lt"` respectively. Does the run time scale as you'd expect?

```{r}
library("Matrix")
library("ywang225ST758")

n_vector=c(100,316,1000,3162,10000)
k_vector=c(2,4,8)

time_matrix=matrix(data=numeric(length(n_vector)*length(k_vector)),
                   nrow = length(k_vector),ncol=length(n_vector))

colnames(time_matrix)=as.character(n_vector)
rownames(time_matrix)=as.character(k_vector)



for(i in 1:length(n_vector)) 
    for(j in 1:length(k_vector))
{
 
  n=n_vector[i]; k=k_vector[j];
  y=1:n;
  
  diags <- list(1:n)
  R_temp <- bandSparse(n, k = -0:k, diag = rep(diags[1],k+1), symm=FALSE)
  A=t(R_temp)%*%R_temp+2*diag(rep(1,n),nrow=n,ncol = n)
  
  R=chol_banded(A,k)
  print(paste0("n=",n," k=",k))
  time=system.time(solve(a=R,b=solve(a=t(R),b=y,system="L"),system="Lt"))
  time_matrix[j,i]=time[1] # i for n; j for k
  print(time[1])
    }


j=1
plot(n_vector,time_matrix[j,],xlab="n",ylab="time",
     main="Relation Between time and n,k in step 5")
lines(n_vector,time_matrix[j,])
j=2
  points(n_vector,time_matrix[j,])
  lines(n_vector,time_matrix[j,],col="red")
j=3
  points(n_vector,time_matrix[j,])
  lines(n_vector,time_matrix[j,],col="blue") 


legend("topleft",legend = paste0("k=",k_vector),col=c("black","red","blue"),
       lty=rep(1,3) )

```



